Stochastic Optimal Control of Single Neuron Spike Trains

### 2014-02-18 draft.tex
Adding ML Note

We need to consider boundary conditions (BCs) for $\v$. If $X(t) = \xth$ then we
have a spike now and $\ts = t$. Thus $$ \v(\xth,t) = (t-\T)^2.$$ At the
threshold, the value function equals the squared difference between the desired
spike time and the realized one. For $x \ll 0$, we assume that $\v$ is not
significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
justified if we impose artificial reflecting boundaries for the diffusion at
$\xmin$. So we need to determine for which $\xmin$ is a reflecting boundary
consistent with the dynamics. For example, we can take $\xmin$ to be two
standard deviations below the mean of the stationary distribution of the
maximally inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set
$\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
that $\xmin \leq -0.5$.
% Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
% the BCs\ldots oh well.

@@

In \cref{fig:HJB_4regimes_value_control_cuts}, we depict three snapshots in time
of the value function and the control, $\v$ and $\a$, at $t = [0,\T/2, \T]$,
i.e.\ at the beginning, middle and end of the relevant time interval. Then in
\cref{fig:HJB_4regimes_value_surf} we show the entire value function, $\v(x,t)$,
the solution to \cref{eq:OC_LS_HJB_full}, for each of the parameter regimes.
% THE CUTS:
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf}
  \caption[labelInTOC]{Snapshots of the value function, $\v$, on the left and on
  the optimal control, $\a$, on the right for $t$ fixed, at the start (blue),
  mid-point (green) and end-point (red), for each of the four regimes.
  The desired spike time is set to $\T=1.5$ assuming the previous spike was at
  $t=0$, the energy penalty, $\eps = 0.001$. The control bounds are $\a \in
  [-2,2]$. 
  Note that all red value function curves go to zero at the threshold to satisfy
  the upper BC in \cref{eq:OC_LS_HJB_full}.
  Note that the green and blue
  curves in F) are lying on top of each other for all $x$.
  \\
  A,B) Supra-Threshold-Low-Noise
  C,D) Supra-Threshold-High-Noise
  E,F) Sub-Threshold-Low-Noise
  G,H) Sub-Threshold-High-Noise
  }
\label{fig:HJB_4regimes_value_control_cuts}
\end{center}
\end{figure}
% THE SURFS:
\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.99\textwidth]{Figs/HJB/Regimes_valuesurf.pdf}
\caption{The numerical solution for the value function $\v(x,t)$ HJB PDE,
\cref{eq:OC_LS_HJB_full}, for the four different parameter regimes.
The desired spike time is set to $\T=1.5$.
The control bounds are $\a \in [-2,2]$. 
The coloured contours for fixed $t$ plotted in thick correspond to the
curves plotted in the left panels of \cref{fig:HJB_4regimes_value_control_cuts}
(blue: $t=0$, green: $t=t*/2$, red: $t=t*$).
\\
A) Supra-Threshold-Low-Noise
B) Supra-Threshold-High-Noise
C) Sub-Threshold-Low-Noise 
D) Sub-Threshold-High-Noise  }
\label{fig:HJB_4regimes_value_surf}
\end{center}
\end{figure}

@@

Note however that increasing $\eps$ has an important effect on the validity of
the approximation, used to form the terminal conditions for the value function,
$\v$, and the adjoint variable, $p$. We assumed that applying $\amax$ is
optimal for $t>\T$. This is indeed true in the limit $\eps \ra 0$. However, even for a
finite value of $\eps$, this may still be the optimal thing to do. Indeed, in
the original calculations, where $\eps=0.001$, see the panels on the left in
\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
the so-obtained value function {\sl implied} that $\a(x,\T) = \amax$. In
other words, our guess that $\a(x, t) = \amax$ for $t > \T$, is
self-consistent with the so-obtained value function. This is akin to confirming an ansatz. For a higher
value of $\eps$, like $\eps = 0.1$, this is no longer the case. The red curves
on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no longer
pushed up at $\a = \amax$, and as such it is no longer valid, strictly speaking,
to assume that the value function can be obtained by assuming $\a(t)|_{t>\T} =
\amax$ and then calculating the expected remaining time-to-spike. The exact same
conclusion can be drawn from the open-loop control. There the optimal control
always (almost) reaches its maximum, $\amax$ while choosing $\eps = 0.001$,
this is no longer so for the higher $\eps = 0.1$. In fact, as we mentioned
after \cref{eq:valuefun_TCs}, the correct thing to do for higher values of $\eps$ is
to push the calculation interval to some $t^+ > \T$, apply the same terminal
condition at $t^+$ instead of $\T$ and then solve backwards (either for the
closed- or open-loop control). One will be guaranteed to find some $t^+ > \T$
that works since eventually the quadratic term $(t-\T)^2$ will dominate the
energy term in the objective, which is linear in $t$. We do not explore this
further here.
% CLOSED_LOOP EPS: 
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
  \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of
  $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
  (red), for each of the four regimes and different values of $\eps$.
  On the left, we use $\eps = 0.001$ and on the right $\eps = 0.1$.
  The desired spike time is set to
  $\T=1.5$ and the bounds are $\a \in [-2,2]$. 
  A,B)
   Supra-Threshold-Low-Noise 
  C,D) 
   Supra-Threshold-High-Noise 
  E,F)
   Sub-Threshold-Low-Noise 
  G,H)
   Sub-Threshold-High-Noise}
\label{fig:HJB_4regimes_control_different_eps} 
\end{center}
\end{figure}
% OPEN_LOOP EPS:
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
  \caption[labelInTOC]{Effect of $\e$ on the open-loop control.
  The deterministic time-trajectory of $\a(t)$ for each
  parameter regime as function of time, $t \in [0, \T]$.
  On the left, we use $\eps = 0.001$ and on the right $\eps =0.1$.
  The desired spike time is set to $\T=1.5$ and the bounds are $\a \in [-2,2]$.
    A,B)
   Supra-Threshold-Low-Noise
  C,D) 
   Supra-Threshold-High-Noise
  E,F)
   Sub-Threshold-Low-Noise
  G,H)
   Sub-Threshold-High-Noise}
  \label{fig:FBK_Regimes_cs_different_es} 
\end{center}
\end{figure}

>>>>

We need to consider boundary conditions (BCs) for $\v$. If $X(t) = \xth$ then we
have a spike now and $\ts = t$. Thus $$ \v(\xth,t) = (t-\T)^2.$$ At the
threshold, the value function equals the squared difference between the desired
spike time and the realized one. For $x \ll 0$, we assume that $\v$ is not
significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
justified if we impose artificial reflecting boundaries for the diffusion at
$\xmin$. So we need to determine for which $\xmin$ is a reflecting boundary
consistent with the dynamics. For example, we can take $\xmin$ to be two
standard deviations below the mean of the stationary distribution of the
maximally inhibited process, i.e.\ setting $\a=\amin$ in
\cref{eq:current_mu_alpha}. That is we set $\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
that $\xmin \leq -0.5$.
% Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
% the BCs\ldots oh well.

@@

In \cref{fig:HJB_4regimes_value_control_cuts}, we depict three snapshots in time
of the value function and the control, $\v$ and $\a$, at $t = [0,\T/2, \T]$,
i.e.\ at the beginning, middle and end of the relevant time interval. Then in
\cref{fig:HJB_4regimes_value_surf} we show the entire value function, $\v(x,t)$,
the solution to \cref{eq:OC_LS_HJB_full}, for each of the parameter regimes.
% THE CUTS:
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf}
  \caption[labelInTOC]{Snapshots of the value function, $\v$, on the left and on
  the optimal control, $\a$, on the right for $t$ fixed, at the start (blue),
  mid-point (green) and end-point (red), for each of the four regimes.
  The desired spike time is set to $\T=1.5$ assuming the previous spike was at
  $t=0$, the energy penalty, $\eps = 0.001$. The control bounds are $\a \in
  [-2,2]$. 
  Note that all red value function curves go to zero at the threshold to satisfy
  the upper BC in \cref{eq:OC_LS_HJB_full}.
  Note that the green and blue
  curves in D) are lying on top of each other for all $x$.
  \\
  A,B) Supra-Threshold-Low-Noise
  C,D) Supra-Threshold-High-Noise
  E,F) Sub-Threshold-Low-Noise
  G,H) Sub-Threshold-High-Noise
  }
\label{fig:HJB_4regimes_value_control_cuts}
\end{center}
\end{figure}
% THE SURFS:
\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.99\textwidth]{Figs/HJB/Regimes_valuesurf.pdf}
\caption{The numerical solution for the value function $\v(x,t)$ HJB PDE,
\cref{eq:OC_LS_HJB_full}, for the four different parameter regimes.
The desired spike time is set to $\T=1.5$.
The control bounds are $\a \in [-2,2]$. 
The coloured contours for fixed $t$ plotted in thick correspond to the
curves plotted in the left panels of \cref{fig:HJB_4regimes_value_control_cuts}
(blue: $t=0$, green: $t=t*/2$, red: $t=t*$).
\\
A) Supra-Threshold-Low-Noise
B) Supra-Threshold-High-Noise
C) Sub-Threshold-Low-Noise 
D) Sub-Threshold-High-Noise  }
\label{fig:HJB_4regimes_value_surf}
\end{center}
\end{figure}

@@

Note however that increasing $\eps$ has an important effect on the validity of
the approximation, used to form the terminal conditions for the value function,
$\v$, and the adjoint variable, $p$. We assumed that applying $\amax$ is
optimal for $t>\T$. This is indeed true in the limit $\eps \ra 0$. However, even for a
finite value of $\eps$, this may still be the optimal thing to do. Indeed, in
the original calculations, where $\eps=0.001$, see the panels on the left in
\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
the so-obtained value function {\sl implied} that $\a(x,\T) = \amax$. In
other words, our guess that $\a(x, t) = \amax$ for $t > \T$, is
self-consistent with the so-obtained value function. This is akin to confirming an ansatz. For a higher
value of $\eps$, like $\eps = 0.1$, this is no longer the case. The red curves
on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no longer
pushed up at $\a = \amax$, and as such it is no longer valid, strictly speaking,
to assume that the value function can be obtained by assuming $\a(t)|_{t>\T} =
\amax$ and then calculating the expected remaining time-to-spike. The exact same
conclusion can be drawn from the open-loop control. There the optimal control
always (almost) reaches its maximum, $\amax$ while choosing $\eps = 0.001$,
this is no longer so for the higher $\eps = 0.1$. In fact, as we mentioned
after \cref{eq:valuefun_TCs}, the correct thing to do for higher values of $\eps$ is
to push the calculation interval to some $t^+ > \T$, apply the same terminal
condition at $t^+$ instead of $\T$ and then solve backwards (either for the
closed- or open-loop control). One will be guaranteed to find some $t^+ > \T$
that works since eventually the quadratic term $(t-\T)^2$ will dominate the
energy term in the objective, which is linear in $t$. We do not explore this
further here.
% CLOSED_LOOP EPS: 
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
  \caption[labelInTOC]{Effect of energy parameter $\e$ on the closed-loop
  control. Snapshots of $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
  (red), for each of the four regimes and different values of $\eps$.
  On the left, we use $\eps = 0.001$ and on the right $\eps = 0.1$.
  The desired spike time is set to
  $\T=1.5$ and the bounds are $\a \in [-2,2]$. 
  A,B)
   Supra-Threshold-Low-Noise 
  C,D) 
   Supra-Threshold-High-Noise 
  E,F)
   Sub-Threshold-Low-Noise 
  G,H)
   Sub-Threshold-High-Noise}
\label{fig:HJB_4regimes_control_different_eps} 
\end{center}
\end{figure}
% OPEN_LOOP EPS:
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
  \caption[labelInTOC]{Effect of $\e$ on the open-loop control.
  The deterministic time-trajectory of $\a(t)$ for each
  parameter regime as function of time, $t \in [0, \T]$.
  On the left, we use $\eps = 0.001$ and on the right $\eps =0.1$.
  The desired spike time is set to $\T=1.5$ and the bounds are $\a \in [-2,2]$.
    A,B)
   Supra-Threshold-Low-Noise
  C,D) 
   Supra-Threshold-High-Noise
  E,F)
   Sub-Threshold-Low-Noise
  G,H)
   Sub-Threshold-High-Noise}
  \label{fig:FBK_Regimes_cs_different_es} 
\end{center}
\end{figure}
